---
title: "Differential expression"
subtitle: "Territorial vs social sockeye salmon"
date: today
params:
  tissue: liver
execute:
  warning: true
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

This notebook runs the DESeq2 analysis for **one tissue**, chosen by the
`tissue` parameter. It replaces `2_DESeq2_analysis-salmon-liver.Rmd` and
`2_DESeq2_analysis-salmon-gonad.Rmd`, which were 571 and 582 lines differing in
five places (finding E5).

Render both tissues with:

```bash
quarto render tag-seq/code/02-differential-expression.qmd -P tissue:liver
quarto render tag-seq/code/02-differential-expression.qmd -P tissue:gonad
```

Package versions are pinned in `renv.lock`; run `renv::restore()` once on a
fresh clone. Nothing is installed at render time.

## Setup

```{r}
#| label: load_libraries
library(DESeq2)       # DESeqDataSetFromMatrix, DESeq, results, lfcShrink, vst, rlog
library(apeglm)       # lfcShrink(type = "apeglm")
library(ashr)         # lfcShrink(type = "ashr")
library(PCAtools)     # pca, screeplot, biplot, pairsplot, getComponents
library(pheatmap)     # sample-correlation heatmap
library(RColorBrewer) # brewer.pal, for the DEG heatmap palette
library(ggpubr)       # ggexport
library(ggplot2)
library(dplyr)
library(tibble)       # rownames_to_column

source("tag-seq/code/_common.R")

cfg <- tissue_config(params$tissue)
dir.create(cfg$dir, recursive = TRUE, showWarnings = FALSE)
cfg[c("tissue", "prefix", "dir", "exclude")]
```

## Load data

Samples are matched between the count matrix and the treatment table **by name**,
not by column position. `prepDE.py` names columns after the GTF files
(`01B_S43_R1.gtf`); the treatment tables use `B01`. `load_counts()` derives one
from the other and asserts the two sets are identical before aligning them.

The previous code assigned `colnames(cts) <- row.names(trt_list)` positionally
and then "checked" it with `all(colnames(cts) %in% rownames(coldata))`, which is
unconditionally `TRUE` after such a rename (finding R1). The assignment happened
to be correct for the committed matrices — verified — but nothing would have
caught a reordered re-run.

```{r}
#| label: load_data
annotation    <- load_gene_annotation()
feature_table <- annotation[, c("gene", "description")]

dat     <- load_counts(cfg)
cts     <- dat$cts
coldata <- dat$coldata

print(dplyr::count(coldata, trt))
cat("samples:", ncol(cts), " genes:", nrow(cts), "\n")
cat("reference level:", levels(coldata$trt)[1], "\n")
```

::: {.callout-important}
## Gene identifiers were repaired

StringTie row names are `gene-<id>|<id>`, and the identifier after the pipe is a
`LOC` number for genes with no assigned symbol and a real symbol for the rest.
The `.Rmd` split each name on the literal string `LOC` and kept the third piece,
which works for `LOC115144855` and turns `gene-arhgap8|arhgap8` into the string
`LOCNA` — for **4,731 genes, 12.5% of the annotation**, and disproportionately
the well-characterised ones.

Because `data.frame()` de-duplicates row names, those genes did not disappear:
they reached the published tables as `LOCNA.1`, `LOCNA.2292` and so on —
identifiers that correspond to nothing, join to nothing, and were therefore
dropped from the gene tables as "no description". In the committed results that
was 3 of 31 significant liver genes and **309 of 1,630 significant gonad genes**.

Identifiers are now taken from after the pipe, which is unique across all 37,942
genes, and descriptions come from the assembly feature table, which covers 37,929
of them and agrees with the old LOCID table on all 33,210 they share.
:::

## Fit the model

```{r}
#| label: deseq
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData   = coldata,
                              design    = ~ trt)
dds <- DESeq(dds)
resultsNames(dds)
```

`social` is set as the reference level explicitly in `load_counts()`, so
`coef = 2` below is `trt_territorial_vs_social` by construction rather than by
alphabetical accident (finding R3). Asserted rather than assumed:

```{r}
#| label: check_contrast
stopifnot(identical(resultsNames(dds)[2], "trt_territorial_vs_social"))
```

Genes are kept if at least a third of samples have 10 or more counts. Note this
filtering happens *after* `DESeq()`, so dispersions were estimated on the full
matrix — preserved from the original, which the committed results depend on.

```{r}
#| label: filter
n_before <- nrow(dds)
dds <- dds[rowSums(DESeq2::counts(dds) >= 10) >= ncol(cts) / 3, ]
cat("genes before filtering:", n_before, "\n")
cat("genes after filtering: ", nrow(dds), "\n")
```

## Ordination and sample clustering

```{r}
#| label: pca
#| fig-width: 8
#| fig-height: 7
p <- pca(assay(vst(dds)), metadata = coldata)

screeplot(p, axisLabSize = 18, titleLabSize = 22)

bp1 <- biplot(p, x = "PC1", y = "PC2",
              colby    = "trt",
              colkey   = c(territorial = "royalblue1", social = "red3"),
              shape    = "trt",
              shapekey = c(territorial = 19, social = 17),
              pointSize = 4,
              showLoadings = FALSE,
              lab = NULL,
              drawConnectors = FALSE,
              ellipse = TRUE, ellipseLevel = 0.95, ellipseFill = FALSE,
              ellipseAlpha = 1/4, ellipseLineSize = 1,
              xlim = cfg$pca_xlim, ylim = c(-100, 100),
              gridlines.major = FALSE, gridlines.minor = FALSE,
              borderWidth = 1,
              legendPosition = "top", legendLabSize = 16, legendIconSize = 6.0) +
  theme(axis.text.x  = element_text(size = 16, colour = "black"),
        axis.text.y  = element_text(size = 16, colour = "black"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.y = element_line(colour = "black"))
bp1

bp2 <- pairsplot(p,
                 components = getComponents(p, c(1:5)),
                 triangle = TRUE, trianglelabSize = 12,
                 hline = 0, vline = 0,
                 pointSize = 1,
                 gridlines.major = FALSE, gridlines.minor = FALSE,
                 colby = "trt",
                 title = "Pairs plot", plotaxes = FALSE,
                 margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), "cm"))
bp2

rld     <- rlog(dds, blind = TRUE)
rld_cor <- cor(assay(rld))
bp3     <- pheatmap(rld_cor, annotation = coldata[, 1:2])
bp3
```

Exports are written to full paths. The original wrapped these in
`setwd("DESEQ_output/<tissue>")` followed by two `setwd('..')` calls, which
knitr resets at every chunk boundary (finding R7).

```{r}
#| label: export_ordination
ggexport(filename = out_path(cfg, cfg$prefix, "-PCA.png"),
         plot = bp1, res = 600, width = 4000, height = 5000)

ggexport(filename = out_path(cfg, cfg$prefix, "-PAIRS.png"),
         plot = bp2, res = 600, width = 6000, height = 6000)

ggexport(filename = out_path(cfg, cfg$prefix, "-pheatmap.png"),
         plot = bp3, res = 600, width = 6000, height = 6000)
```

## Contrasts and shrinkage

```{r}
#| label: contrasts
res_table        <- results(dds, contrast = c("trt", "territorial", "social"), alpha = 0.05)
res_table_normal <- lfcShrink(dds, coef = 2, type = "normal")
res_table_apeglm <- lfcShrink(dds, coef = 2, type = "apeglm")
res_table_ashr   <- lfcShrink(dds, coef = 2, type = "ashr")
```

```{r}
#| label: ma_plots
#| fig-width: 10
#| fig-height: 4
draw_ma <- function() {
  par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
  xlim <- c(1, 1e5); ylim <- c(-3, 3)
  for (nm in c("normal", "apeglm", "ashr")) {
    DESeq2::plotMA(get(paste0("res_table_", nm)), xlim = xlim, ylim = ylim, main = nm, cex = .8)
    abline(h = c(-1.5, 1.5), col = "dodgerblue", lwd = 2)
  }
}
draw_ma()

pdf(file = out_path(cfg, "MA_plots.pdf"))
draw_ma()
invisible(dev.off())
```

## Results tables

Significance is `padj < 0.05`, and only that.

The previous code also defined `lfc.cutoff <- 1.5` and `svalue.cutoff <- 0.005`,
never used them, and then labelled its outputs
`DEG_apeglm-s0.005_lfc1.5` — describing a filter that never ran (finding R2).
The unused variables are gone and the labels now say what was actually done. The
counts themselves are unchanged.

```{r}
#| label: tidy_tables
padj.cutoff <- 0.05

as_gene_tibble <- function(x) {
  x %>% data.frame() %>% rownames_to_column(var = "gene") %>% as_tibble()
}

res_table        <- as_gene_tibble(res_table)
res_table_normal <- as_gene_tibble(res_table_normal)
res_table_apeglm <- as_gene_tibble(res_table_apeglm)
res_table_ashr   <- as_gene_tibble(res_table_ashr)

cat("Total genes:", nrow(res_table), "\n")
cat("Genes after NAs removed:", nrow(na.omit(res_table)), "\n")

res_table        <- na.omit(res_table)
res_table_normal <- na.omit(res_table_normal)
res_table_apeglm <- na.omit(res_table_apeglm)
res_table_ashr   <- na.omit(res_table_ashr)
```

```{r}
#| label: annotate
annotate_genes <- function(tbl) {
  tbl %>%
    left_join(feature_table, by = "gene") %>%
    arrange(-log2FoldChange) %>%
    mutate(treatment = cfg$tissue)
}

all_genes        <- annotate_genes(res_table)
all_genes_normal <- annotate_genes(res_table_normal)
all_genes_apeglm <- annotate_genes(res_table_apeglm)
all_genes_ashr   <- annotate_genes(res_table_ashr)

sig_genes        <- annotate_genes(filter(res_table,        padj < padj.cutoff))
sig_genes_normal <- annotate_genes(filter(res_table_normal, padj < padj.cutoff))
sig_genes_apeglm <- annotate_genes(filter(res_table_apeglm, padj < padj.cutoff))
sig_genes_ashr   <- annotate_genes(filter(res_table_ashr,   padj < padj.cutoff))
```

```{r}
#| label: single_gene_counts
## Normalised counts for every significant gene, per shrinkage estimator.
single_gene_counts <- function(sig) {
  if (nrow(sig) == 0) return(data.frame())
  do.call(rbind, lapply(sig$gene, function(locid) {
    d <- plotCounts(dds, gene = locid, intgroup = "trt", returnData = TRUE)
    d$gene   <- locid
    d$sample <- rownames(d)
    rownames(d) <- NULL
    d
  }))
}

counts_unshrunk <- single_gene_counts(sig_genes)
counts_normal   <- single_gene_counts(sig_genes_normal)
counts_apeglm   <- single_gene_counts(sig_genes_apeglm)
counts_ashr     <- single_gene_counts(sig_genes_ashr)

write.table(counts_unshrunk, out_path(cfg, cfg$prefix, "-unshrunk-single-gene-counts.csv"), row.names = FALSE, sep = ",")
write.table(counts_normal,   out_path(cfg, cfg$prefix, "-normal-single-gene-counts.csv"),   row.names = FALSE, sep = ",")
write.table(counts_apeglm,   out_path(cfg, cfg$prefix, "-apeglm-single-gene-counts.csv"),   row.names = FALSE, sep = ",")
write.table(counts_ashr,     out_path(cfg, cfg$prefix, "-ashr-single-gene-counts.csv"),     row.names = FALSE, sep = ",")
```

```{r}
#| label: write_tables
gene_counts <- data.frame(count = c(
  nrow(cts), nrow(dds), nrow(cts) - nrow(dds),
  nrow(all_genes), nrow(all_genes_normal), nrow(all_genes_apeglm), nrow(all_genes_ashr),
  nrow(sig_genes), nrow(sig_genes_normal), nrow(sig_genes_apeglm), nrow(sig_genes_ashr)
), row.names = c(
  "genes_before_filtering", "genes_after_filtering", "genes_dropped",
  "DEGs_all-genes", "DEGs_all-genes-normal", "DEGs_all-genes-apeglm", "DEGs_all-genes-ashr",
  "DEG_unshrunken-p0.05", "DEG_normal-p0.05", "DEG_apeglm-p0.05", "DEG_ashr-p0.05"
))
gene_counts

write.table(all_genes,        out_path(cfg, cfg$prefix, "-ALL-DEG.csv"),            row.names = FALSE, sep = ",")
write.table(all_genes_normal, out_path(cfg, cfg$prefix, "-ALL-DEG-normal.csv"),     row.names = FALSE, sep = ",")
write.table(all_genes_apeglm, out_path(cfg, cfg$prefix, "-ALL-DEG-apeglm.csv"),     row.names = FALSE, sep = ",")
write.table(all_genes_ashr,   out_path(cfg, cfg$prefix, "-ALL-DEG-ashr.csv"),       row.names = FALSE, sep = ",")
write.table(sig_genes,        out_path(cfg, cfg$prefix, "-SIG-DEG-unshrunken.csv"), row.names = FALSE, sep = ",")
write.table(sig_genes_normal, out_path(cfg, cfg$prefix, "-SIG-DEG-normal.csv"),     row.names = FALSE, sep = ",")
write.table(sig_genes_apeglm, out_path(cfg, cfg$prefix, "-SIG-DEG-apeglm.csv"),     row.names = FALSE, sep = ",")
write.table(sig_genes_ashr,   out_path(cfg, cfg$prefix, "-SIG-DEG-ashr.csv"),       row.names = FALSE, sep = ",")
write.table(gene_counts,      out_path(cfg, cfg$prefix, "-gene-counts.csv"),        row.names = TRUE, sep = ",")
```

## Volcano plots

```{r}
#| label: volcanoes
#| fig-width: 6
#| fig-height: 4
volcano <- function(tbl) {
  ggplot(tbl, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = log2FoldChange > 0))
}

volcano_specs <- list(
  list(file = "Volcano_all_genes.png",        data = all_genes),
  list(file = "Volcano_all_genes_normal.png", data = all_genes_normal),
  list(file = "Volcano_all_genes_ashr.png",   data = all_genes_ashr),
  list(file = "Volcano_all_genes_apeglm.png", data = all_genes_apeglm),
  list(file = "Volcano_sig_genes.png",        data = sig_genes),
  list(file = "Volcano_sig_genes_normal.png", data = sig_genes_normal),
  list(file = "Volcano_sig_genes_ashr.png",   data = sig_genes_ashr),
  list(file = "Volcano_sig_genes_apeglm.png", data = sig_genes_apeglm)
)

for (spec in volcano_specs) {
  p <- volcano(spec$data)
  print(p)
  ggsave(out_path(cfg, spec$file), plot = p,
         dpi = 600, device = "png", width = 6, height = 4, units = "in")
}
```

## Expression heatmap

Column annotation colours are derived from `coldata`. The previous code wrote
`rep("royalblue1", 15)` / `rep("red3", 15)` for liver and `14` / `14` for gonad,
hardcoding both the group sizes and the column order (finding R4).

```{r}
#| label: heatmap
sigs <- res_table_apeglm %>%
  filter(padj < 0.05) %>%
  left_join(feature_table, by = "gene")

df_sigs <- as.data.frame(sigs)
stopifnot(!anyDuplicated(df_sigs$gene))
rownames(df_sigs) <- df_sigs$gene

rlog_out <- rlog(dds, blind = FALSE)
mat_sigs <- assay(rlog_out)[rownames(df_sigs), rownames(coldata)]
colnames(mat_sigs) <- rownames(coldata)

## Z-score each gene across samples.
mat_scaled_sigs <- t(apply(mat_sigs, 1, scale))
colnames(mat_scaled_sigs) <- colnames(mat_sigs)

side_colours <- trt_side_colours(coldata)
stopifnot(length(side_colours) == ncol(mat_scaled_sigs))

png(filename = out_path(cfg, cfg$heatmap_prefix, "_heatmap_pval_genes_apeglm.png"),
    res = 1200, width = 6000, height = 6000, units = "px", pointsize = 6)
heatmap(mat_scaled_sigs, scale = "none",
        col = colorRampPalette(brewer.pal(11, "RdYlBu"))(256),
        ColSideColors = side_colours)
invisible(dev.off())
```

## Session

```{r}
#| label: session
sessionInfo()
```
