---
title: "GO enrichment"
subtitle: "goseq against the NCBI annotation for GCF_006149115.2"
date: today
---

Replaces the DAVID web-tool step that produced `DAVID_GOterms.txt` and
`DAVID_KEGG_pathway.txt`, for which no script, input list, background or query
date was recorded (both now in `archive/david-2023/`, see the README there). It also replaces `X_GOterm_Annotation.Rmd` and
`X_GOSeq_annotation.Rmd`, which belonged to a different experiment entirely
(finding B3).

**Annotation source.** NCBI publishes a GO annotation for this exact assembly,
vendored into the repository so the analysis does not depend on NCBI staying
reachable:

```
tag-seq/genome/GCF_006149115.2_Oner_1.1_gene_ontology.gaf.gz
```

GAF 2.2, generated by NCBI 2023-12-02. No BLAST, no web tool, no manual step.

**KEGG is available after all.** An earlier check of the KEGG REST organism
list came back empty and was misread as "no such organism". KEGG does carry
*Oncorhynchus nerka* as organism `one` (genome T07952, 37,942 proteins — the
same gene count as our matrix). Its gene identifiers are NCBI GeneIDs, so they
join to our `LOC` ids directly. The two REST responses are vendored alongside
the GAF; see `tag-seq/genome/kegg_one_SOURCE.txt` for the URLs and retrieval
date. No ortholog transfer is needed.

```{r}
#| label: setup
library(goseq)
library(GO.db)
library(AnnotationDbi)
library(dplyr)
source("tag-seq/code/_common.R")

GO_DIR <- "tag-seq/GO_output"
dir.create(GO_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Gene-to-GO mapping

```{r}
#| label: gaf
gaf_path <- file.path(PATHS$genome, "GCF_006149115.2_Oner_1.1_gene_ontology.gaf.gz")

gaf <- read.delim(gaf_path, comment.char = "!", header = FALSE,
                  quote = "", stringsAsFactors = FALSE)
names(gaf)[c(2, 3, 4, 5, 7, 9)] <-
  c("GeneID", "symbol", "qualifier", "GO", "evidence", "aspect")

## GAF qualifiers beginning NOT record an explicit *absence* of a function.
gaf <- gaf %>% filter(!grepl("^NOT", qualifier))

gene2cat <- gaf %>% select(gene = symbol, GO) %>% distinct()

cat("annotations:", nrow(gene2cat), " genes:", dplyr::n_distinct(gene2cat$gene),
    " GO terms:", dplyr::n_distinct(gene2cat$GO), "\n")
print(table(gaf$aspect))
```

## Gene lengths, and why they are not used to weight the test

`goseq` was written for whole-transcript RNA-seq, where longer genes accumulate
more reads and are easier to call differentially expressed, and its default
Wallenius test re-weights every gene by a spline fitted to that relationship.
This analysis uses the plain **hypergeometric** test instead, for three reasons
that came out of the first clean-machine render (PR #4, 2026-09-06):

1. **TagSeq does not have that bias.** It sequences one fragment from the 3′ end
   of each transcript, so read count does not scale with transcript length, and
   the premise of the correction does not hold for this assay.
2. **The length association here is strong and in the opposite direction.**
   In gonad, 27% of the shortest-decile genes are differentially expressed
   against about 5% in every other decile (Wilcoxon p ≈ 1e-119). The Wallenius
   weighting therefore down-weighted short genes heavily, and the pathways it
   suppressed were the ones made of short genes, oxidative phosphorylation
   first among them (59 of 188 genes DE: Wallenius p 2e-5, hypergeometric
   p 1e-18).
3. **It was not reproducible.** The spline fitted by `nullp()` is numerically
   platform-sensitive, and the Wallenius results sat at the threshold: the same
   inputs gave FDR 0.004 / 0.039 / 0.045 for the three enriched terms on macOS
   and 0.060 / 0.057 / 0.081 on Linux, so `check-reproduction.R` failed on the
   gonad enrichment tables and nothing else. The hypergeometric test is exact
   and deterministic.

`nullp()` is still run, because `goseq()` takes its gene universe from the
object it returns and because its length-bias plot is the evidence for point 2.
Under `method = "Hypergeometric"` the fitted weights are not used.

The plan was to take lengths from `sequences/GCF_006149115.2_Oner_1.1_mRNA.gff`,
but that file was tracked at **zero bytes** — the `bedtools` step that should
have written it (documented in `01-upstream-alignment.qmd`, "Generate mRNA
feature track") never produced anything that was committed. It has since been
untracked rather than left in place as an empty file asserting that it exists
(finding E7). Lengths come instead from the assembly feature table, which is
committed and complete: the median mRNA interval length per gene, keyed by
`GeneID`, which is our `LOC` identifier without the prefix.

```{r}
#| label: lengths
annotation   <- load_gene_annotation()
gene_lengths <- setNames(annotation$length, annotation$gene)
geneid2gene  <- setNames(annotation$gene, as.character(annotation$GeneID))

cat("genes with a length:", sum(!is.na(gene_lengths)), "\n")
```

## KEGG pathway mapping

```{r}
#| label: kegg
kegg_links <- read.delim(file.path(PATHS$genome, "kegg_one_gene2pathway.tsv"),
                         header = FALSE, stringsAsFactors = FALSE,
                         col.names = c("gene", "pathway"))
kegg_names <- read.delim(file.path(PATHS$genome, "kegg_one_pathways.tsv"),
                         header = FALSE, stringsAsFactors = FALSE,
                         col.names = c("pathway", "name"))

## KEGG keys genes by NCBI GeneID ("one:115101128"). Map through the feature
## table rather than pasting a "LOC" prefix on: that would be right for
## LOC-numbered genes and wrong for every symbol-named one.
kegg2cat <- kegg_links %>%
  transmute(gene    = unname(geneid2gene[sub("^one:", "", gene)]),
            pathway = sub("^path:", "", pathway)) %>%
  filter(!is.na(gene)) %>%
  distinct()

kegg_names$name <- sub(" - Oncorhynchus nerka \\(sockeye salmon\\)$", "", kegg_names$name)

cat("gene-pathway links:", nrow(kegg2cat),
    " genes:", dplyr::n_distinct(kegg2cat$gene),
    " pathways:", dplyr::n_distinct(kegg2cat$pathway), "\n")
```

## Enrichment

The universe is every gene DESeq2 returned an adjusted p-value for, and the
differentially expressed set is those with `padj < 0.05` — the same rule the
results tables use. The test is the hypergeometric over-representation test,
Benjamini–Hochberg corrected across the categories tested.

```{r}
#| label: enrich
run_goseq <- function(tissue, cat_map, label, term_lookup) {
  cfg <- tissue_config(tissue)
  all_deg <- read.csv(out_path(cfg, cfg$prefix, "-ALL-DEG-apeglm.csv"),
                      stringsAsFactors = FALSE)

  ## goseq cannot fit a length bias for genes with no length.
  usable <- all_deg %>% filter(gene %in% names(gene_lengths))

  de <- as.integer(usable$padj < 0.05)
  names(de) <- usable$gene

  cat(sprintf("\n== %s / %s ==\n  universe %d (of %d tested; %d lack a length)\n  DE %d\n  universe genes with a category: %d\n",
              tissue, label, length(de), nrow(all_deg), nrow(all_deg) - nrow(usable),
              sum(de), sum(names(de) %in% cat_map[[1]])))

  if (sum(de) < 2) {
    cat("  too few DE genes to test\n")
    return(NULL)
  }

  pwf <- nullp(de, bias.data = gene_lengths[names(de)], plot.fit = TRUE)
  title(main = paste(tissue, "-", label, "- proportion DE by transcript length"))

  ## Hypergeometric, not the default Wallenius: see "Gene lengths" above.
  res <- goseq(pwf, gene2cat = as.data.frame(cat_map), method = "Hypergeometric")

  res <- res %>%
    mutate(FDR = p.adjust(over_represented_pvalue, method = "BH")) %>%
    term_lookup() %>%
    arrange(over_represented_pvalue)

  dir.create(file.path(GO_DIR, tissue), recursive = TRUE, showWarnings = FALSE)
  write.csv(res, file.path(GO_DIR, tissue, sprintf("%s-%s-all.csv", tissue, label)),
            row.names = FALSE)
  sig <- res %>% filter(FDR < 0.05)
  write.csv(sig, file.path(GO_DIR, tissue, sprintf("%s-%s-enriched-FDR05.csv", tissue, label)),
            row.names = FALSE)

  cat(sprintf("  categories tested %d | FDR < 0.05: %d\n", nrow(res), nrow(sig)))
  list(tissue = tissue, label = label, res = res, sig = sig)
}

go_terms <- function(res) {
  res %>%
    mutate(term     = AnnotationDbi::Term(category),
           ontology = AnnotationDbi::Ontology(category)) %>%
    select(category, term, ontology, over_represented_pvalue, FDR, numDEInCat, numInCat)
}

kegg_terms <- function(res) {
  res %>%
    left_join(kegg_names, by = c("category" = "pathway")) %>%
    select(category, name, over_represented_pvalue, FDR, numDEInCat, numInCat)
}

results <- list()
for (tis in c("liver", "gonad")) {
  results[[paste0(tis, "-GO")]]   <- run_goseq(tis, gene2cat, "GO",   go_terms)
  results[[paste0(tis, "-KEGG")]] <- run_goseq(tis, kegg2cat, "KEGG", kegg_terms)
}
```

## Enriched terms

```{r}
#| label: report
for (r in results) {
  if (is.null(r)) next
  cat("\n===", r$tissue, r$label, "- top 12 by p-value  (significant at FDR<0.05:",
      nrow(r$sig), ") ===\n")
  print(head(as.data.frame(r$res), 12), row.names = FALSE)
}
```

```{r}
#| label: session
sessionInfo()
```
